Recalibrating Gravitational Wave Phenomenological Waveform Model

### 2023-05-17 src/tex/ms.tex
intrin_space fix

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right ($\chi_1>0>\chi_2$) 
regions respond to optimization differently. In the top-left 
region, waveforms generally improve with along optimization. However, mismatches 
in the bottom-right region do not improve after optimization. Most waveforms even 
turned worse after optimization.

@@

\begin{table}[t]
	\centering
	\begin{tabularx}{0.8\columnwidth}{@{\extracolsep{\fill}}lrrr}
		\toprule\midrule Code         & $q$ & $\chi_1$ & $\chi_2$ \\
		\midrule\midrule SXS:BBH:0172 & 1.0 & 0.98     & 0.98     \\
		SXS:BBH:0152 & 1.0 & 0.60     & 0.60     \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1417 & 4.0 & 0.40     & 0.50     \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1426 & 8.0 & 0.48     & 0.75     \\
		SXS:BBH:0167 & 8.0 & 0.00     & 0.00     \\ \midrule 
		SXS:BBH:0370 & 1.0 & -0.20    & 0.40     \\
		SXS:BBH:2092 & 1.0 & -0.50    & 0.50     \\
		SXS:BBH:0330 & 1.0 & -0.80    & 0.80     \\
		SXS:BBH:2116 & 2.0 & -0.30    & 0.30     \\
		SXS:BBH:2111 & 2.0 & -0.60    & 0.60     \\
		SXS:BBH:0335 & 2.0 & -0.80    & 0.80     \\
		SXS:BBH:0263 & 3.0 & -0.60    & 0.60     \\
		SXS:BBH:2133 & 3.0 & -0.73    & 0.85     \\
		SXS:BBH:0263 & 4.0 & -0.80    & 0.80     \\ \midrule 
		SXS:BBH:0156 & 1.0 & -0.95    & -0.95    \\
		SXS:BBH:0151 & 1.0 & -0.60    & -0.60    \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1418 & 4.0 & -0.40    & -0.50    \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1419 & 8.0 & -0.80    & -0.80    \\
		SXS:BBH:0063 & 8.0 &  0.00    &  0.00    \\ \midrule 
		SXS:BBH:0304 & 1.0 & 0.50     & -0.50    \\
		SXS:BBH:0327 & 1.0 & 0.80     & -0.80    \\
		SXS:BBH:2123 & 2.0 & 0.30     & -0.30    \\
		SXS:BBH:2128 & 2.0 & 0.60     & -0.60    \\
		SXS:BBH:2132 & 2.0 & 0.87     & -0.85    \\
		SXS:BBH:2153 & 3.0 & 0.30     & -0.30    \\
		SXS:BBH:0045 & 3.0 & 0.50     & -0.50    \\
		SXS:BBH:0292 & 3.0 & 0.73     & -0.85    \\ \midrule\bottomrule
	\end{tabularx}
	\caption{List of waveforms used in recalibrating coefficients in 4 regions. From top to down are the top-right ($\chi_1,\chi_2>0$), top-left ($\chi_1<0<\chi_2$), bottom-left ($\chi_1,\chi_2<0$) and bottom-right ($\chi_1>0>\chi_2$) regions. Note that for the top-right and bottom-left regions, waveforms are chosen to have $\chi_1\approx\chi_2$, while the training waveforms for the other two regions are chosen to have $\chi_1\approx-\chi_2$.}
	\label{tab:quadrants}
\end{table}

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.~\ref{fig:ps_q148}). The
training waveforms used for fitting are listed in Table~\ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\leq4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe
improvements in waveform mismatch for all waveforms, except those in the bottom-right 
region. For equal-spin waveforms lying in the proximity of the diagonal axis,
the mismatch has additional improvement due to most training waveforms being
equal-spin. In the top-left region, the improvement is significant, except for a
few defects caused by some testing waveforms having $q>4$. This can be seen clearer in 
Fig.~\ref{fig:all_quadrants}, which we see that the optimized histogram shifts downward uniformly. 
The optimized waveforms in the bottom-right region have a higher mismatch than the original ones,
indicating poor fitting. These waveforms correspond to the waveforms in the high mismatch tail 
in Fig.~\ref{fig:q148}, which we can see in Fig.~\ref{fig:all_quadrants}. Thus, we show that 
further optimization in a smaller subspace does not improve the results. \kw{I am a bit confused about why you can get worse result.}

@@

We have shown the result of recalibrating waveform coefficients. One thing to
note is that our recalibration procedure is not exactly the same as the original
calibration. For instance, we use a different set of NR waveforms, frequency
range, etc. Nonetheless, as the decrease in mismatch is rather significant, this
optimization procedure should be able to improve the accuracy of IMRPhenomD on a
similar scale regardless of the differences. Here, this result serves as a
demonstration of the general method used.  
% In Fig.~\ref{fig:0154}, the error in the inspiral region is around halved for
% both amplitude and phase. Similarly, for the merger-ringdown region, the
% optimized waveforms also show great improvement.

@@

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top left
region than the bottom right along a degeneracy line. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec. \ref{sec:result}. However, even with separate optimizations for each
regions, Fig.\ref{fig:ps_q148_quadrant} indicates that the bottom-right region
exhibits similar mismatches as before, while the top-left region's performance
has improved. This observation suggests that the ansatz is specific to certain
regions of the parameter space, with a preference for BBH events featuring
$\chi_1<0$ and $\chi_2>0$.

@@

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple} or IMRPhenomX \citep{pratten2020setting} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. It will be interesting 
to recalibratethe IMRPhenomX model \citep{pratten2020setting}. Because it is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models or EOB models introduced previously. Such an approach could 
simplify NR waveform calibration procedures and lead to the improvement of 
existing models.

@@

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the bottom-right 
region cannot be improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
we can see that this recalibration process gives further
improvement to the top-left region while the bottom-right region shows similar
result. This indicates that the ansatz does not fit waveforms in this region. 
This phenomenon is due to the reduced spin approximation used in parameterizing the 
ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.

>>>>

Since we know intrinsic parameters play an important role in the ansatz, we
would like to investigate how intrinsic parameters affect the recalibration
process. First, we plot the parameter space of $q$ vs. $\chi_{\mathrm{PN}}$ in
Fig. \ref{fig:ps_q148_qchi}. We can see near non-spinning waveforms demonstrate
more consistent improvement, since the ansatz are developed base on non-spinning 
waveforms. Also, the original coefficients were fitted using NR waveforms with
equal or similar spin, hence the model prefers waveforms with similar spin. On 
the other hand, spinning waveforms can either improve or worsen in terms of 
mismatch base on their intrinsic parameters. To further discuss the behavior of 
non-spinning waveforms, we plot the parameter space of $\chi_1$ vs. $\chi_2$ in
Fig.~\ref{fig:ps_q148}. Waveforms along the diagonal axis, i.e. 
$\chi_1\approx\chi_2$, show good mismatch improvements as discussed above. 
Meanwhile, the top-left ($\chi_1<0<\chi_2$) and bottom-right ($\chi_1>0>\chi_2$) 
regions respond to optimization differently. In the top-left 
region, waveforms generally improve slightly with along optimization. However, waveforms 
in the bottom-right region do not improve after optimization. Some waveforms even 
turned worse after optimization.

@@

\begin{table}[t]
	\centering
	\begin{tabularx}{0.8\columnwidth}{@{\extracolsep{\fill}}lrrr}
		\toprule\midrule Code         & $q$ & $\chi_1$ & $\chi_2$ \\
		\midrule\midrule SXS:BBH:0172 & 1.0 & 0.98     & 0.98     \\
		SXS:BBH:0152 & 1.0 & 0.60     & 0.60     \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1417 & 4.0 & 0.40     & 0.50     \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1426 & 8.0 & 0.48     & 0.75     \\
		SXS:BBH:0063 & 8.0 & 0.00     & 0.00     \\ \midrule 
		SXS:BBH:0370 & 1.0 & -0.20    & 0.40     \\
		SXS:BBH:2092 & 1.0 & -0.50    & 0.50     \\
		SXS:BBH:0330 & 1.0 & -0.80    & 0.80     \\
		SXS:BBH:2116 & 2.0 & -0.30    & 0.30     \\
		SXS:BBH:2111 & 2.0 & -0.60    & 0.60     \\
		SXS:BBH:0335 & 2.0 & -0.80    & 0.80     \\
		SXS:BBH:0263 & 3.0 & -0.60    & 0.60     \\
		SXS:BBH:2133 & 3.0 & -0.73    & 0.85     \\
		SXS:BBH:0263 & 4.0 & -0.80    & 0.80     \\ \midrule 
		SXS:BBH:0156 & 1.0 & -0.95    & -0.95    \\
		SXS:BBH:0151 & 1.0 & -0.60    & -0.60    \\
		SXS:BBH:0001 & 1.0 & 0.00     & 0.00     \\
		SXS:BBH:1418 & 4.0 & -0.40    & -0.50    \\
		SXS:BBH:0167 & 4.0 & 0.00     & 0.00     \\
		SXS:BBH:1419 & 8.0 & -0.80    & -0.80    \\
		SXS:BBH:0063 & 8.0 &  0.00    &  0.00    \\ \midrule 
		SXS:BBH:0304 & 1.0 & 0.50     & -0.50    \\
		SXS:BBH:0327 & 1.0 & 0.80     & -0.80    \\
		SXS:BBH:2123 & 2.0 & 0.30     & -0.30    \\
		SXS:BBH:2128 & 2.0 & 0.60     & -0.60    \\
		SXS:BBH:2132 & 2.0 & 0.87     & -0.85    \\
		SXS:BBH:2153 & 3.0 & 0.30     & -0.30    \\
		SXS:BBH:0045 & 3.0 & 0.50     & -0.50    \\
		SXS:BBH:0292 & 3.0 & 0.73     & -0.85    \\ \midrule\bottomrule
	\end{tabularx}
	\caption{List of waveforms used in recalibrating coefficients in 4 regions. From top to down are the top-right ($\chi_1,\chi_2>0$), top-left ($\chi_1<0<\chi_2$), bottom-left ($\chi_1,\chi_2<0$) and bottom-right ($\chi_1>0>\chi_2$) regions. Note that for the top-right and bottom-left regions, waveforms are chosen to have $\chi_1\approx\chi_2$, while the training waveforms for the other two regions are chosen to have $\chi_1\approx-\chi_2$.}
	\label{tab:quadrants}
\end{table}

We divided the parameter space into four regions to analyze the effect of the
recalibration procedure on each region separately (Fig.~\ref{fig:ps_q148_quadrant}). The
training waveforms used for fitting are listed in Table~\ref{tab:quadrants}. The
top-left and bottom-right regions have limited data for $q>4$, hence the results are
only valid up to $q\lesssim4$. From Fig.~\ref{fig:ps_q148_quadrant}, we observe that 
waveforms lying in the lower diagonal have mismatches worsened while other waveforms 
generally improve. For the top-right and bottom-left regions, we see that the optimization prefers waveforms with $\chi_1<\chi_2$. This further  In the top-left region, the improvement is significant, except for a
few defects caused by some testing waveforms having $q>4$. This can be seen clearer in 
Fig.~\ref{fig:all_quadrants}, which we see that the optimized histogram shifts downward 
rather uniformly. The optimized waveforms in the bottom-right region have two distinct 
features. Within this region, waveforms in the bottom-right corner improves significantly 
while others worsened. This phenomenon can also be seen in Fig.~\ref{fig:all_quadrants}, 
where the mismatch distribution shifted upward. Hence, we can deduce that the loss function 
descended into a different minimum compared to Fig.~\ref{fig:ps_q148}, making the optimized 
coefficients not useful in GW analysis. THus, further optimizing in a smaller region does 
not generally improve the results.
%\kw{I am a bit confused about why you can get worse result.}

@@

We have shown the result of recalibrating waveform coefficients. One thing to
note is that our recalibration procedure is not exactly the same as the original
calibration. For instance, we use a different set of NR waveforms, frequency
range, etc. Nonetheless, as the decrease in mismatch is rather significant, this
optimization procedure should be able to improve the accuracy of IMRPhenomD on a
similar scale regardless of the differences. Here, the result serves as a
demonstration of the general method used.  
% In Fig.~\ref{fig:0154}, the error in the inspiral region is around halved for
% both amplitude and phase. Similarly, for the merger-ringdown region, the
% optimized waveforms also show great improvement.

@@

The reduced spin approximation is a major contributor to the inaccuracies
observed in the ansatz. In IMRPhenomD, this approximation employs a single spin
parameter, $\chi_{\mathrm{PN}}$, as described in Sec. \ref{sec:method}. The
parameterization of BBH mergers using a single spin parameter results in a
degeneracy within the parameter space. Specifically, black hole events with
different spins may generate the same waveform due to identical values of
$\chi_{\mathrm{PN}}$, leading to erroneous results, particularly for highly
unequal spin events. This degeneracy produces straight lines in the parameter
space with negative slopes that depend on the mass ratio, as shown in
Fig.\ref{fig:ps_q148}. Notably, the ansatz performs better in the top-left
region than the bottom-right region. In an attempt to address
this issue, we partitioned the parameter space into four regions, as described
in Sec. \ref{sec:result}. With separate optimizations for each
regions, Fig.\ref{fig:ps_q148_quadrant} indicates that the top-left region's 
performance has improved, while the bottom-right region exhibits two distinct 
features. This is because the loss function has descended into a different 
minimum which highly favors events with $\chi_1\approx1$ and $\chi_2\approx-1$. 
This observation suggests the ansatz is specific to certain regions of the 
parameter space, with a preference for BBH events lying in the upper diagonal 
region, and it hardly fits with waveforms lying in the lower diagonal region.

@@

Although our study primarily focused on the IMRPhenomD model, this simple yet
versatile approach can be applied to other differentiable GW models, such as the
IMRPhenomP \citep{hannam2014simple, khan2019phenomenological} or IMRPhenomX \citep{pratten2020setting,pratten2021computationally} 
models within the same family. By jointly optimizing a new set of coefficients,
it is expected that both models can be enhanced since they share similar
construction principles to the IMRPhenomD model. For instance, they also use PN
approximants as part of the ansatz in the inspiral segment. It will be interesting 
to recalibratethe IMRPhenomXAS model \citep{pratten2020setting}. Because it is
parameterized by an additional anti-symmetric spin parameter, it is expected not
to exhibit the degeneracy previously described. With the currently developing 
{\jax} IMRPhenomXAS model in {\ripple}, A more detailed investigation
may provide valuable insights into the systematics of the Phenom models.
Furthermore, this approach is applicable to other GW model families, such as NR
surrogate models or EOB models introduced in Sec. \ref{sec:intro}. Such an approach could 
simplify NR waveform calibration procedures and lead to the improvement of 
existing models.

@@

Furthermore, we investigated how the intrinsic parameters affect the
improvement. Fig.~\ref{fig:ps_q148} shows that the optimization procedure has a
certain preference for waveforms lying in the top-left region while the bottom-right 
region hardly improved. To further test this result, we recalibrate
waveforms in separate regions in parameter space. From Fig.~\ref{fig:ps_q148_quadrant}, 
we can see that this recalibration process gives further
improvement to the top-left region while the bottom-right region falls into a different 
loss minimum. This indicates that the ansatz does not fit certain waveforms in this 
region, hence introduces bias to downstream GW analysis. This phenomenon is mainly 
due to the reduced spin approximation used in parameterizing the 
ansatz, where degeneracies between $\chi_1$ and $\chi_2$ are introduced.
